---
title: "Setting up a Southern Ocean ecosystem model using mizer"
author: Julia L. Blanchard
date: Feb, 5, 2020
place: Horbart
output:
  pdf_document: default
  html_document: default
---


# Introduction

Here, we will push the boundaries of size spectrum theory by introducing species that may not necessarily fit within  the conceptualisaion of the trait-based and size spectrum models. To do so we will think about how different assumptions would need to change to accomodate species beyond fish- such as large baleen whales, seabirds and other species.

To begin with we will read in some data for predators in the Kerguelen regions of the Southern Ocean.

```{r}
#get required packages
#library(devtools)
#most up to date master branch of mizer
#install_github("sizespectrum/mizer")
#install_github("astaaudzi/mizer-rewiring", ref = "temp-model-comp")
#documentation here:
#https://sizespectrum.org/mizer/dev/index.html
library(mizer)
require(tidyverse)
# 
# remotes::install_github("sizespectrum/mizerExperimental")
# library(mizerExperimental)

```

### Data
Let's read in some biomasses of predators in the Southern Ocean.

```{r}

#read in data for prydz bay
dat<-read.csv("data/pred_group_biomasses.csv",header=T)

#read groups for southern ocean model
#groups<-readRDS("data/groups.RDS")
#write.csv(groups,"data/groups.csv")

groups <-read.csv("data/groups2.csv")
groups<-merge(groups,dat,by.x="species",by.y="group",all=T)

#have a look at plot
plot(groups$w_inf,groups$biomass.tperkm2,xlab="log Maximum Weight [g]", ylab=" log Total Biomass", log="xy",col="blue",pch=16)
text(groups$w_inf,groups$biomass.tperkm2,labels=groups$species,cex=0.5)

# could plot the paramter allometries here to explore
```


## Set-up mizer model

Next let's read in the parameters we've put together for a Southern Ocean (Prydz Bay) ecosystem.

```{r}
#read groups again for southern ocean model, this time subsetting key groups
groups<-read.csv("data/groups2.csv")[-1,]

#groups <- readRDS("data/groups.RDS")[-1,]
#check fails these tests:
groups$w_mat25 >= groups$w_mat 
groups$w_mat25 <= groups$w_min
groups$w_inf <= groups$w_mat25

#[-1,]
#fix one value 
#groups[7,"w_min"]<-4000
# read interaction matrix
# get the interaction matrix
theta <- readRDS("data/prydz_inter.RDS")[-1,-1]
#[-1,-1]

 
```

We will pass these parameters to mizer to set up a new multispecies model.

```{r}
 
mf.params <- newMultispeciesParams(species_params=groups,interaction=theta,kappa = 1e3,min_w_pp = 1e-16,w_pp_cutoff = 1,r_pp = 10,z0pre = 0.01)

# mf.params@species_params$metab[1] <-0.01
mf.params@species_params$h[] <- 10*mf.params@species_params$ks[] 

# mf.params@species_params$gamma[1] <- 1000*(mf.params@species_params$gamma[1])
mf.params@species_params$alpha[] <- 0.6
#mf.params@species_params$z0[1] <- 0.01

#higher mean betas
 mf.params@species_params$beta[mf.params@species_params$species == "large.divers"] <-6000
 mf.params@species_params$beta[mf.params@species_params$species == "apex.predators"] <-1000
# # wider feeding kernels
 mf.params@species_params$sigma[] <- 2
 mf.params@species_params$sigma[mf.params@species_params$species == "baleen.whales"] <- 4


# mf.params@species_params$w_mat25[1] <- 0.005*mf.params@species_params$w_inf[1]
mf.params@species_params$erepro <- ifelse(mf.params@species_params$erepro==1,0.001,mf.params@species_params$erepro)

# mf.params@interaction[] <-0.1
#mf.params@interaction[,1] <-0
#mf.params@interaction[1,1] <-1

#mf.params@species_params$interaction_p[2:11] <-0.5

# hand tuning to gte in line with data
mf.params@species_params$R_max <- 10*mf.params@species_params$R_max
#mf.params@species_params$R_max <- log10(1000*params@resource_params$kappa*params@species_params$w_inf^-1) 


# needs to be very different for marine mammals? use different repro assumptions (fixed offspring density per year)?



# mf.params@species_params$w_mat <- 0.5*mf.params@species_params$w_inf
# mf.params@species_params$w_mat[mf.params@species_params$species == "flying.birds"] <- 0.75*mf.params@species_params$w_inf[mf.params@species_params$species == "flying.birds"]
# mf.params@species_params$w_mat[mf.params@species_params$species == "small.divers"] <- 0.75*mf.params@species_params$w_inf[mf.params@species_params$species == "small.divers"]
# mf.params@species_params$w_mat[mf.params@species_params$species == "medium.divers"] <- 0.75*mf.params@species_params$w_inf[mf.params@species_params$species == "medium.divers"]
# mf.params@species_params$w_mat[10:12] <- 0.75*mf.params@species_params$w_inf[10:12]
# setParams(mf.params)

mf.params<- setParams(mf.params) # take the new paramewters and change kappa


sim <- project(mf.params, t_max=100,dt = 0.1)
plot(sim)
#plotlyGrowthCurves(sim,species="macrozooplankton")
plotlyFeedingLevel(sim)
# feeding level satioation for some groups, except for the seabirds
# macrozooplankton - they are not growing enough,why?
#tuneParams(mf.params)
plotlyGrowthCurves(sim,percentage = T)
plotlySpectra(sim,power = 2)
```

If we combine the biomassess from the model and add the data to  plot we cna look at how close/far we are.

```{r}
# get biomass and covert to tonnes per area
plotBioData <- function (sim,dat){
simbio<-getBiomass(sim)
simbio<-data.frame(species=mf.params@species_params$species,w_inf=mf.params@species_params$w_inf,simbio=simbio[100,])

simbio<-as.data.frame(merge(simbio,dat,by.x="species",by.y="group",all=T))

#have a look at plot with both modelled total biomass and observed for top predators
plot(simbio$w_inf,simbio$simbio,xlab="log Maximum Weight [g]", ylab=" log Total Biomass", log="xy",col="grey",
pch=16,ylim=c(min(simbio$simbio,simbio$biomass.tperkm2,na.rm=T),max(simbio$simbio,simbio$biomass.tperkm2,na.rm=T)))
points(simbio$w_inf,simbio$biomass.tperkm2,col="blue",pch=1)
text(simbio$w_inf,simbio$simbio,labels=simbio$species,cex=0.5)
return(simbio)
}

plotBioData(sim,dat)


# 
# 
# ## have a look at paramters - metabolism
# plot(mf.params@species_params$w_inf,mf.params@species_params$ks,log="x")
# text(mf.params@species_params$w_inf,mf.params@species_params$ks,labels=params@species_params$species,cex=0.5)
# 
# ## have a look at paramters - h
# plot(mf.params@species_params$w_inf,mf.params@species_params$h,log="x")
# text(mf.params@species_params$w_inf,mf.params@species_params$h,labels=params@species_params$species,cex=0.5)
# # 

 # plot(mf.params@species_params$w_inf,mf.params@species_params$ks/mf.params@species_params$h,log="x")
 # text(mf.params@species_params$w_inf,mf.params@species_params$ks/mf.params@species_params$h,labels=params@species_params$species,cex=0.5)
# 

# # ## have a look at paramters - gamma
#  plot(mf.params@species_params$w_inf,mf.params@species_params$gamma,log="xy")
#  text(mf.params@species_params$w_inf,mf.params@species_params$gamma,labels=params@species_params$species,cex=0.5)
# # 
# # 
# # ## have a look at paramters - wmat/winf
#  plot(mf.params@species_params$w_inf,mf.params@species_params$w_mat/mf.params@species_params$w_inf,log="xy",xlab=c("Weight"),y=c("Wmat/Winf"))
#  text(mf.params@species_params$w_inf,mf.params@species_params$w_mat/mf.params@species_params$w_inf,labels=params@species_params$species,cex=0.5)
 
# changes need to maturation sizes - zoop groups -> wmat 50% of winf
```


# Model Assumptions
### Feeding kernels
Let's change some of the parameters which make some of these groups less fishy! The feeding kernel in teh above model is set up for fish. Here we use the model of Heneghan to describe a more planktivours feeding kernel form zooplankton and baleen whales.
```{r}

####### feeding kernels - use box kernels for planktovores instead?
params<-mf.params

#new predation kernel varies with pred size and species but is the same for all prey species, need to fill this in with values, according to Ryans' code
pred_kern <- array(0, dim=c(length(params@species_params$species),length(params@w),length(params@w_full)),dimnames=list(params@species_params$species,params@w,params@w_full))

#### change the feeding kernel
pred_kern[]<- mf.params@species_params$beta

#mvals<-groups$m ### need to enter these manually as removed from file because of new m parameter in mizer setparams function
#shift over microzoop
#mvals[1]<--2

mvals <-c(
NA,
0.73,
NA,
NA,
NA,
NA,
NA,
NA,
NA,
NA,
1.5)


### Ryan's feeding kernel:
for (i in 1:length(mvals)) {
  
  D.z <- 2*(3*params@w*1e12/(4*pi))^(1/3) # convert body mass g to ESD (um)
  betas =  (exp(0.02*log(D.z)^2 - mvals[i] + 1.832))^3
  
  if (!is.na(mvals[i])) pred_kern[i,,]<- betas
  
}
  pred_kern[]<-exp(-0.5*sweep(log(sweep(sweep(pred_kern,3,params@w_full,"*")^-1,2,params@w,"*")),1,params@species_params$sigma,"/")^2)
  pred_kern[] <- sweep(pred_kern,c(2,3),combn(params@w_full,1,function(x,w)x<params@w,w=params@w),"*") # find out the untrues and then multiply
  
##### box feeding kernel for baleen whales  

  
  
  
####### psi function
  
#   # Change allocation to reproduction = check for endotherms/determinate growers (should be 1 at wmat not winf)
#  psi_r=20
#  params@psi[] <- unlist(tapply(params@w,1:length(params@w),function(wx,Winf,Wmat,n){
#    ((1 + (wx/(Wmat))^-psi_r)^-1) * (wx/Winf)^(1-n)},Winf=params@species_params$w_inf,Wmat=params@species_params$w_mat,n=params@n))
# # # Set w < 1% of Wmat to 0
#  params@psi[unlist(tapply(params@w,1:length(params@w),function(wx,Wmat)wx<(Wmat*0.01),Wmat=params@species_params$w_mat))] <- 0
# # # Set all m > M to 1 # Check this is right...
#  params@psi[unlist(tapply(params@w,1:length(params@w),function(wx,Winf)(wx/Winf)>1,Winf=params@species_params$w_inf))] <- 1
#  
# ########## erepro
#  
#  #and erepro - should this decline with Winf or shoudl it just be a very small number 0.005?
# #should this actually be closer to 1 for mammals?
params@species_params$erepro <-0.05*params@species_params$w_inf^(-0.75)
# # need to think about how this sounds in writing
# 
# ######### juvenile mortality
# 
# ## add declining intraspecific juvenile mortality to background mortality 
 for (i in 1: length(params@species_params$species)) params@mu_b[i,] <- params@mu_b[i,]  + 0.1*params@w^-0.25
# 


```

Let's take a look now.
```{r}
  
params<- setParams(params, pred_kernel = pred_kern,kappa=1e4)

# run again, intialised from the final state above
new.mf<-project(params,t_max=100,initial_n=sim@n[100,,],initial_n_pp=sim@n_pp[100,])

#plot(new.mf)

# run again for longer, from the final state above
#new.mf<-project(params,t_max=500,initial_n=new.mf@n[100,,],initial_n_pp=new.mf@n_pp[100,])
   
plot(new.mf)  
plotlyFeedingLevel(new.mf)  
plotlyGrowthCurves(new.mf,percentage = T) 
plotlyBiomass(new.mf,percentage = T) 
plotlySpectra(new.mf,power=2) 
plotBioData(new.mf,dat)

```

That helps baleen whales - and impact macrozoooplankton, but some of the predator groups are still not getting enough food.

Let's follow the protocol in "toyexample1" and tune kappa to see if we can achive a better size spectrum. We can use Rshiny:

```{r}
library(shiny)
#sim <- new.mf
runApp("data/shiny-equilibrium")
# is there a way to save the final chosen values?
```



### Prey size selection

What range of values should the seabirds have - should we use a box kernel for those groups? I seem to remember there are distinct size ranges for the different seabirds.

### Metabolic costs

Are these correct for the different groups? What about the maximum intake rate coefficients - h? Do these sclae in the way that is expected for endotherms?

### Maturation and reproduction

We need to check the assumptions about maturation and reproduction for the non-fish groups.

The marine mammals and seabirds show stop growth once they reach maturation size. Need to make sure some functioanl groups are not mixtures of these type of taxa. Therefore matruation size and asymptotic size are very close. They should probably alos follow a sharp transition to maturation size, as they follow determinate growth.

Density dependence: Erepro, Rmax.

Costs of reproduction for marine mammals and seabirds? epreor can be used for this.
Assume constant reprodcution 1 chick/calf per yer per couple - reuiqres change to getERepro, or hard wire via assumptions of constant reporoduction - 1/2*sum(population density)



Should we change the reproduction function to not use the stock-recruitmnet assumption? There are other options now:https://sizespectrum.org/mizer/reference/index.html#section-density-dependent-reproduction


### Starvation mortality

Should we include startvation morality ? This would free up competition for resources....

Need starvation mortality for marine mammal and seabirds especially?

### Other sources of mortality

Currently assuming a fairly low backgournd "other" moratlity that is constant within fucntioan group but declines with aymptotic size (Brown et al. 2004). Sensescence mortality has been used in other work, but is tricky to parameterise. Though this might be needed for squid or other life histories that dies after reproduction.

### Plankton dynamics

Currently we use the default plankton dynamics and parameter values. What should these be for the Southern Ocean? Do we have any information from regional shelf sea biogeochmical models? The previous information of satellite data seem limited. 

So, we decided to ask the question: What does the plankton need to be to feed the rest of the size spectrum (at equlibrium)? Does anyone have size spectrum slopes and intercepts for this system? (Angus Atkinson?)

However, we could also incude stochasticity...more work though example here: https://rpubs.com/gustav/plankton-anchovy

Samik Datta and I have also looked at seasonality - hopefully will build extension, but how important for this?

# Simple Model Calibration

Note:"toyexample1" for more  info on this method.

Rather than continue to manually adjust and tune the parameters we will try to automatically calibrate the model with data and estimate our uncertain parameters. We are particularly uncertian about:  Rmax( this will influence the biomass of each group). Here, we will use an inverse calibraion where we have population level biomass density data for 6 of the marine mammal and seabird groups (obtained from the inputs for an EcoPath model). 

This might take AWHILE. Go watch some Netflix.

```{r}
# change kappa and erepro based on shiny exploration
#  params@species_params$erepro[] <-10^-4
params <- setParams(params,kappa=10^4)
  
# subset row for these species only:
subset<- c("flying.birds","small.divers","medium.divers","apex.predators","large.divers","baleen.whales")

ind<-which(params@species_params$species %in%  subset ==T)

# define the initial parameters to send to optimisation code below

# we need initlal values for species Rmaxs, log10 scale
vary<-log10(params@species_params$R_max)[ind] 

#my magic equation:
vary <- log10(1000*params@resource_params$kappa*params@species_params$w_inf^-0.75)[ind]
# or use completley made up values, same for each species test for effects of initial values

#groups<-groups[groups$species %in%  subset,]

#the following getError function returns the least sum of squared errors between the observations and model estimates
preddat = dat[dat$group %in%  subset,]

getError <- function(vary=vals,params=params,dat=preddat,data_type="biomass",timetorun=100,whichspecies=ind) {
  params@species_params$R_max[whichspecies]<-10^vary
  sim <- project(params, effort = 0, t_max = timetorun, dt=0.1)
          
          ## what kind of data and output do we have?
         
          if (data_type=="biomass") {
         
          output <-getBiomass(sim)[timetorun,]   # units:1 g/m2 = 1 t/km2
          
          simbio<-data.frame(species=mf.params@species_params$species,w_inf=mf.params@species_params$w_inf,simbio=output) 
          
          simbio<-as.data.frame(merge(simbio,dat,by.x="species",by.y="group",all=T))
           
          output <- simbio[!is.na(simbio[,4]),3]
          
          obs<-simbio[!is.na(simbio[,4]),4]
         }
         
          if (data_type=="SSB") {
          output <-getSSB(sim)[timetorun,]   #could change to getBiomass if appropriate, also check units.
          }
          
          if (data_type=="catch") {
         output <-getYield(sim)[timetorun,] #### CHECK UNITS !! grams per year? the data are in tonnes per year so converting to tonnes.
          }
          
  pred <- log(output)
  obs  <- log(obs)

  # sum of squared errors, here on log-scale of predictions and data (could change this or use other error or likelihood options)
  
   discrep <- pred - obs

   discrep <- (sum(discrep^2))
  
  # can use a strong penalty on the error to ensure we reach a minimum of 10% of the data (biomass or catch) for each species
 #  if(any(pred < 0.1*dat)) discrep <- discrep + 1e10
  
    return(discrep)

   }

## test it


err<-getError(vary,params, dat = preddat,data_type="biomass")
#test
err
```



```{r}
# this time carry out optimisation, using optim(), with pred biomasses only

# could estimate erepro instead, and hold R_max constant

vals<-optim(par=vary,getError,params=params,method ="L-BFGS-B",lower=c(rep(1e-2,6)),upper= c(rep(10,6)))


#vals<-optim(par=vary,getError,params=params,method ="L-BFGS-B",lower=c(rep(3,12)),upper= c(rep(20,12)))

# plug back into model
# make sure kappa and erepro are the same
#params@species_params$erepro[] <-1

params <- setParams(params,kappa=1e4)
# optim values:
params@species_params$R_max[ind] <- 10^vals$par 
# set the param object
params<-setParams(params)
sim <- project(params, effort = 0, t_max = 100, dt=0.1)
plot(sim)

# save vals - may want to repeat this setp depending on diagnostics, after changing some parameters
saveRDS(vals,"optim_vals3.RDS")

# save params - may want to repeat this setp depending on diagnostics, after changing some parameters
saveRDS(params,"optim_param3.RDS")
saveRDS(sim,"optim_sim3.RDS")


```



#### Step 4. Check that the physiological recruitment, \RDI, is much higher than the realised recruitment, \RDD. High RDI/RDD ratio. NOt sure this holds for no-fish groups though.

```{r}
params<- readRDS("optim_param3.RDS")
sim <- readRDS("optim_sim3.RDS")
getRDI(params)/getRDD(params)

```


#### Step 5. Verify the model after the above step by comparing the model with data. 

Eg. species biomass or abundance distrubtions, feeding level, naturality mortality, growth, vulnerablity to fishing (fmsy) and catch, diet composition.. Many handy functions for plotting these are available here: https://sizespectrum.org/mizer/reference/index.html


```{r}

## check obs vs pred
pred_bio <-melt(getBiomass(sim)[100,])
pred_bio$species <- row.names(pred_bio)
pred_bio<-as.data.frame(merge(pred_bio,preddat,by.x="species",by.y="group",all=T))
           
p <- ggplot() + # plot predicted and observed yields
        geom_point(data = pred_bio[pred_bio$species %in%  subset,], 
            aes(x = log10(value), y = log10(biomass.tperkm2), color = species)) +
   # plot optimal fit line
        geom_abline(color = "black", slope = 1, intercept = 0) + 
  xlab("log10 Predicted Biomass Density") + 
  ylab("log10 Observed Biomass Density") +
  xlim(log10(0.001), log10(2)) +
  ylim(log10(0.001), log10(2)) 
p
 
# check other plots

plotlyGrowthCurves(sim,percentage = T) 
plotlyFeedingLevel(sim) 
plotGrowthCurves(sim,"baleen.whales") # check each species 
plotlySpectra(sim,power=2)
plotDiet(params,"baleen.whales") 
plotBioData(sim,dat)
# What would happen if changed parameters above based on different assumptions? Need to check all of these in detail to  make sure they sense!
```
